Liveness-Based Garbage Collection for Lazy Languages

### 2015-05-13 paper/paper.tex
liveness

A  notion  related  to  liveness  is  {\em  demand}.   While  liveness
represents the  future use of variables, demands  represent the future
use of the  values of expressions. The demand on  an expression $e$ is
again a  set of  access paths---that subset  of $\{\acar,\acdr\}^\ast$
which the context of $e$ may explore of $e$'s result.  To see the need
for  demands,  consider   the  let  expression  $\pi:(\LET~x\leftarrow
(\CDR~y)~\IN~  \pi':\RETURN~x)$.  Assume  that  the  context  of  this
expression produces  the liveness $[x \mapsto  \lbrace \epsilon, \acar
  \rbrace]$ at $\pi'$.  Due to the \LET\ definition which binds $x$ to
$(\CDR~y)$, the  liveness of $x$ at  $\pi'$ now becomes  the demand on
$(\CDR~y)$.    This,  in   turn,  generates   the   liveness  $\lbrace
y.\epsilon,  y.\acdr,  y.\acdr\acar  \rbrace$  at $\pi$  (the  set  of
$y$-rooted  accesses  required  to  explore $\lbrace  \epsilon,  \acar
\rbrace$  paths of  the result  of  $(\CDR~y)$)\footnote{The classical
  analogue  is strong  liveness analysis,  where  $y$ and  $z$ have  a
  liveness of  $\lbrace\epsilon\rbrace$ at  the entry $n:  x:=y+z$, if
  and only if  $x$ has a liveness of  $\lbrace\epsilon\rbrace$ at exit
  of $n$.   In our terminology, the  liveness of $x$ at  exit from $n$
  becomes the  demand on $y+z$.  This, in turn,  generates the  liveness $\lbrace
  y.\epsilon, z.\epsilon \rbrace$ at the entry of $n$.}.
%% Exploring $\lbrace \epsilon, \acar \rbrace$ of the
%% result of $(\CDR~y)$ first requires access using $y$ ( and the \CDR\ field of the cons cell pointed by
%% $y$.
Note that  liveness refers to  access using variables and  fields, and
not  to $\CONS$  cells (i.e.\  to edges  in the  memory graph,  not to
locations themselves).
\begin{figure*}[t!]
  \begin{tabular}{cc}
    \begin{boxedminipage}{.51\textwidth}
        \small
        \renewcommand{\arraystretch}{1}{
          \begin{uprogram}
            \UNL{1} (\DEFINE\ (\length~\xl)
            \UNL{2}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\ (\NULLQ~\xl) \IN
            \UNL{3} \hspace*{.05cm} $\pi_2\!\!:\,$(\SIF\ $\psi_1\!\!:\,$ \px
            \UNL{4} \hspace*{.27cm} $\psi_2\!\!:\, (\RETURN ~0)$
            \UNL{4} \hspace*{.29cm}    $\pi_3\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\xl)  \IN
            \UNL{5} \hspace*{.34cm}   $\pi_4\!\!:\, $(\LET~\py\  $\leftarrow$  (\length~\pu)  \IN
            \UNL{6} \hspace*{.34cm} $\pi_5\!\!:\,
            $(\LET~\pz\ $\leftarrow$ (1~+~\py) \IN
            \UNL{7} \hspace*{.34cm} $\psi_3\!\!:\, (\RETURN ~\pz)$))))))
        \end{uprogram}}
        \renewcommand{\arraystretch}{1}{
	  \begin{uprogram}
	  \UNL{1} $\pi_\mainpgm\!\!:\, \pi_6\!\!:\, $(\LET\  \pa\  $\leftarrow$ \ldots  \IN
	  \UNL{2}   \hspace*{.78cm}   $\pi_7\!\!:\,      $ (\LET\ \pb\  $\leftarrow$  \ldots \IN 
          \UNL{3}   \hspace*{.78cm}    $\pi_8\!\!:\,
          $(\LET\ \pw\  $\leftarrow$  (\length\ \pb) \IN
          \UNL{4}  \hspace*{.78cm}  $\psi_{4}\!\!:\,$ (\RETURN\ \pw)))))))
\end{uprogram}}
        %%} 
    \end{boxedminipage}

@@

        \small
        \begin{tabular}{lc}
        $\cal{M_{\mbox{\tiny \length}}} = $ 
            \begin{tabular}{lc}            
              $[~ep_1 \mapsto \lbrace \px.\sigma_{\length}, \xl.\cbar\sigma_{\length} \rbrace ,~ep_2 \mapsto \lbrace \rbrace , $\\
              $~~ep_3 \mapsto \lbrace \pz.\sigma_{\length},
                \py.\cbar\sigma_{\length}, \pu.\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}, $\\
              $~~~~~~~~~~~~~~\;\;
                \xl.\acdr\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}
                \rbrace~\rbrack$ \\
            \end{tabular} \\
            $\cal{P_{\mbox{\tiny \length}}} = $
              \begin{tabular}{lc}
                $~[
                  ~\pi_1 \mapsto \lbrace ep_1, ep_2, ep_3 \rbrace, $ \
                $ \pi_2 \mapsto \lbrace ep_1, ep_2, ep_3
                  \rbrace, $\\
                $~~\psi_1 \mapsto \lbrace ep_1 \rbrace, $ \
                $ \psi_2 \mapsto \lbrace ep_2 \rbrace, $ \
                $ \pi_5 \mapsto \lbrace ep_3 \rbrace, $ \\
                $~~\pi_6 \mapsto \lbrace ep_3 \rbrace, $ \
                $ \pi_7 \mapsto \lbrace ep_3 \rbrace,~ \psi_3 \mapsto \lbrace ep_3 \rbrace~\rbrack$
              \end{tabular}\\
              $\cal{M_{\mbox{\tiny \main}}} = $ 
              \begin{tabular}{lc}            
                $[$
                  $ep_4 \mapsto \lbrace \pw.\sigma_{\mathit{\!all}}$, $\pb.\Lf{\length}{\mbox{1}}{\sigma_{\mathit{\!all}}}$, $\ldots  \rbrace $ \
                  $]$
              \end{tabular} \\
              $\cal{P_{\mbox{\tiny \main}}} = $
                \begin{tabular}{lc}
                  $~\lbrack$~$\pi_6 \mapsto \lbrace ep_4 \rbrace, $ \
                  $ \pi_7 \mapsto \lbrace ep_4 \rbrace, $ \
                  $\pi_8 \mapsto \lbrace ep_4 \rbrace,~\psi_4 \mapsto \lbrace ep_4 \rbrace~\rbrack$  
                \end{tabular}
        \end{tabular}

@@

We next  describe the function $\mathcal{L}$  that propagates liveness
across  expressions.   This brings  us  to  the  second reason why 
liveness analysis for lazy and eager languages are different---the
point  where  an  expression   gets  evaluated  cannot  be  statically
determined.   %% To deal  with this,  we introduce  an analysis  which is
%% flow-insensitive  but path-sensitive\comment{Checked  the terminology,
%%   path-sensitive means certain prooperties  holding along a path only,
%%   http://en.wikipedia.org/wiki/Data-flow\_analysis} into our analysis.
To see  the consequences of this, consider  an analysis  of the body  of $\length$  with a
demand $\sigma$.  Clearly, this is also the demand on the return value
$z$   giving  the   liveness  at   the  program   point   $\psi_3$  as
$\{\pz.\sigma\}$.  The liveness at $\psi_3$ is propagated backwards as
in traditional liveness analysis.

@@

Since   \pz\  is   $(1  +   \py)$,  the   liveness  of   $\sigma$  for
\pz\ translates into a demand  of $\sigma$ on $(1+\py)$, which in turn
generates a  liveness of $\{\py.\epsilon\}$ at  $\pi_5$.  We now make
the following observations:
\begin{enumerate}
\item While the liveness of \py\  should be killed at $\pi_4$ since it
  is  beyond the scope  of the  definition of  \py, we  do not  do so.
  However, this  does not cause any  imprecision in the way  of the GC
  marking more cells as live, since \py\ is yet to come into existence
  at $\pi_4$,  and its liveness will  not be consulted  by the garbage
  collector at  this point. 
\item In an eager language, \py\ would not have been live at $\psi_3$,
  a program point  beyond its last use.  In  a lazy language, however,
  the  definition $\pz  \leftarrow  (1+\py)$, does  not  result in  an
  evaluation  of  $(1+\py)$;  instead   a  closure  is  created.   The
  evaluation  actually takes  place while  evaluating $(\RETURN~\pz)$.
  We therefore  record  the liveness of $\py$ at the program  
  point $\psi_3$,  indeed at every  program point from $\psi_3$  up to
  the beginning  of the program (since  we do not  kill liveness).
\end{enumerate}

We thus regard  the body of the function as a  tree and consider paths
from  the  root  of this  tree  to  program  points which  trigger  an
evaluation\footnote{Program  points which  trigger  evaluation  are
  labelled with  $\psi$ {\color {red} ---slightly late in the day}}---\SIF\ conditions and  \RETURN\ expressions.
We call such paths {\em e-paths}.  For the function \length, there are
three such paths $ep_1 = \langle \pi_1, \pi_2, \psi_1\rangle$, $ep_2 =
\langle\pi_1, \pi_2, \psi_2\rangle$ and  $ep_3 = \langle \pi_1, \pi_2,
\pi_3, \pi_4, \pi_5, \psi_3\rangle$.

Given  a function  $\mathit{f}$ and  a demand  $\sigma$,  the liveness
analysis   comes   the    map   $\cal{M}_{\mathit   f}^{\sigma}$   (or
$\cal{M}_{\mathit  f}$ when  the  demand $\sigma$  is  clear from  the
context) maps  a e-path to a  liveness environment arising  out of all
uses of variables  along the path. A second  map $\cal{P}_{\mathit f}$
maps a  program point to the set  of e-paths sharing the  point.  If a
program point  happens to fall  on more than  one such path,  then the
liveness environment  of the program point is  the variable-wise union
of the liveness environments of  the individual paths.  In the example
program,  since ${\mathcal P}_{\length}(\pi_1)  = \lbrace  ep_1, ep_2,
ep_3\rbrace$,  the liveness  at $\pi_1$  is  ${\cal M_{\length}}(ep_1)
\cup {\cal  M_{\length}}(ep_2)\cup {\cal M_{\length}}(ep_3)$.   In the
sequel   we   shall  describe   how   to   obtain  $\cal{M}$   through
$\mathcal{L}$.  The  formulation of  $\cal{P}$ is easy  and we  do not
elaborate it any further.

@@

The  expression $\psi:(\RETURN~x)$  with  a demand  $\sigma$ forces  an

@@

is created and added to  the union.  Also notice the consequences
of  laziness---the entire  expression including  the condition  is not

>>>>

A  notion  related  to  liveness  is  {\em  demand}.   While  liveness
represents the  future use of variables, demands  represent the future
use of the  values of expressions. The demand on  an expression $e$ is
again a  set of  access paths---that subset  of $\{\acar,\acdr\}^\ast$
which the context of $e$ may explore of $e$'s result.  To see the need
for  demands,  consider   the  let  expression  $\pi:(\LET~x\leftarrow
(\CDR~y)~\IN~  \pi':\RETURN~x)$.  Assume  that  the  context  of  this
expression produces  the liveness $[x \mapsto  \lbrace \epsilon, \acar
  \rbrace]$ at $\pi'$.  Due to the \LET\ definition which binds $x$ to
$(\CDR~y)$, the  liveness of $x$ at  $\pi'$ now becomes  the demand on
$(\CDR~y)$.    This,  in   turn,  generates   the   liveness  $\lbrace
y.\epsilon,  y.\acdr,  y.\acdr\acar  \rbrace$  at $\pi$  (the  set  of
$y$-rooted  accesses  required  to  explore $\lbrace  \epsilon,  \acar
\rbrace$  paths of  the result  of  $(\CDR~y)$)\footnote{The classical
  analogue  is strong  liveness analysis,  where  $y$ and  $z$ have  a
  liveness of  $\lbrace\epsilon\rbrace$ at  the entry $n:  x:=y+z$, if
  and only if  $x$ has a liveness of  $\lbrace\epsilon\rbrace$ at exit
  of $n$.   In our terminology, the  liveness of $x$ at  exit from $n$
  becomes the  demand on $y+z$.  This, in turn,  generates the  liveness $\lbrace
  y.\epsilon, z.\epsilon \rbrace$ at the entry of $n$.}.
%% Exploring $\lbrace \epsilon, \acar \rbrace$ of the
%% result of $(\CDR~y)$ first requires access using $y$ ( and the \CDR\ field of the cons cell pointed by
%% $y$.
Note that  liveness refers to  access using variables and  fields, and
not  to $\CONS$  cells (i.e.\  to edges  in the  memory graph,  not to
locations themselves). 
\begin{figure*}[t!]
  \begin{tabular}{cc}
    \begin{boxedminipage}{.51\textwidth}
        \small
        \renewcommand{\arraystretch}{1}{
          \begin{uprogram}
            \UNL{1} (\DEFINE\ (\length~\xl)
            \UNL{2}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\ (\NULLQ~\xl) \IN
            \UNL{3} \hspace*{.05cm} $\pi_2\!\!:\,$(\SIF\ $\psi_1\!\!:\,$ \px
            \UNL{4} \hspace*{.27cm} $\pi_3\!\!:\,
            $(\LET\ \pv\ $\leftarrow 0$ \IN
            \UNL{5} \hspace*{.32cm} $\pi_4\!\!:\, (\RETURN~\psi_2:\pv)$
            \UNL{4} \hspace*{.29cm}    $\pi_5\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\xl)  \IN
            \UNL{5} \hspace*{.34cm}   $\pi_6\!\!:\, $(\LET~\py\  $\leftarrow$  (\length~\pu)  \IN
            \UNL{6} \hspace*{.34cm} $\pi_7\!\!:\,
            $(\LET~\pz\ $\leftarrow$ (1~+~\py) \IN
            \UNL{7} \hspace*{.34cm} $\pi_8\!\!:\, (\RETURN~\psi_3:\pz)$)))))))
        \end{uprogram}}
        \renewcommand{\arraystretch}{1}{
	  \begin{uprogram}
	  \UNL{1} $(\DEFINE\ (\main)$
          \UNL{2} \!\!$\pi_9\!\!:\, (\LET\  \pa\  \leftarrow$ \ldots  \IN
	  \UNL{3}   \hspace*{.05cm}$\pi_{10}\!\!:\,      $ (\LET\ \pb\  $\leftarrow$  \ldots \IN 
          \UNL{4}   \hspace*{.15cm}    $\pi_{11}\!\!:\,
          $(\LET\ \pw\  $\leftarrow$  (\length\ \pb) \IN
          \UNL{5}  \hspace*{.25cm}  $\pi_{12}\!\!:\, (\RETURN~\psi_4:\pw)))))$
\end{uprogram}}
        %%} 
    \end{boxedminipage}

@@

        \small
        \begin{tabular}{lc}
        $\cal{M_{\mbox{\tiny \length}}} = $ 
            \begin{tabular}{lc}            
              $[~ep_1 \mapsto \lbrace \px.\sigma_{\length}, \xl.\cbar\sigma_{\length} \rbrace ,~ep_2 \mapsto \lbrace \rbrace , $\\
              $~~ep_3 \mapsto \lbrace \pz.\sigma_{\length},
                \py.\cbar\sigma_{\length}, \pu.\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}, $\\
              $~~~~~~~~~~~~~~\;\;
                \xl.\acdr\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}
                \rbrace~\rbrack$ \\
            \end{tabular} \\
            $\cal{P_{\mbox{\tiny \length}}} = $
              \begin{tabular}{lc}
                $~[
                  ~\pi_1 \mapsto \lbrace ep_1, ep_2, ep_3 \rbrace, $ \
                $ \pi_2 \mapsto \lbrace ep_1, ep_2, ep_3
                  \rbrace, $\\
$ ~~~\pi_3 \mapsto \lbrace ep_2
                  \rbrace $, $ \pi_4 \mapsto \lbrace ep_2
                  \rbrace$, $\pi_5 \mapsto \lbrace  ep_3 \rbrace$, \\
                $~~~\pi_6 \mapsto \lbrace ep_3 \rbrace,  \pi_7 \mapsto
                  \lbrace ep_3\rbrace, \pi_8 \mapsto \lbrace ep_3 \rbrace~\rbrack$
              \end{tabular}\\
              $\cal{M_{\mbox{\tiny \main}}} = $ 
              \begin{tabular}{lc}            
                $[$
                  $ep_4 \mapsto \lbrace \pw.\sigma_{\mathit{\!all}}$, $\pb.\Lf{\length}{\mbox{1}}{\sigma_{\mathit{\!all}}}$, $\ldots  \rbrace $ \
                  $]$
              \end{tabular} \\
              $\cal{P_{\mbox{\tiny \main}}} = $
                \begin{tabular}{lc}
                  $~\lbrack$~$\pi_6 \mapsto \lbrace ep_4 \rbrace, $ \
                  $ \pi_7 \mapsto \lbrace ep_4 \rbrace, $ \
                  $\pi_8 \mapsto \lbrace ep_4 \rbrace,~\psi_4 \mapsto \lbrace ep_4 \rbrace~\rbrack$  
                \end{tabular}
        \end{tabular}

@@

We next  describe the function $\mathcal{L}$  that propagates liveness
across expressions.  Note that,  unlike eager languages, the syntactic
occurrence  of an expression  in a  program denotes  the point  of 
creation of its closure, not necessarily its evaluation.
%% To   deal  with   this,   we  introduce   an   analysis  which   is
%%flow-insensitive but path-sensitive\comment{Checked the terminology,
%%path-sensitive means certain properties  holding along a path only,
%%http://en.wikipedia.org/wiki/Data-flow\_analysis} into our analysis.
To see the  consequences of this, consider an analysis  of the body of
$\length$ with a demand $\sigma$.  Clearly, this is also the demand on
the return value $z$ giving the liveness at the program point $\pi_8$
as $\{\pz.\sigma\}$.  The liveness at $\pi_8$ is propagated backwards
as in traditional liveness analysis.

@@

Since   \pz\  is   $(1  +   \py)$,  the   liveness  of   $\sigma$  for
\pz\ translates into a demand  of $\sigma$ on $(1+\py)$, which in turn
generates a  liveness of $\{\py.\epsilon\}$ at  $\pi_7$.  We now make
the following observations:
\begin{enumerate}
\item While the liveness of \py\  should be killed at $\pi_6$ since it
  is  beyond the scope  of the  definition of  \py, we  do not  do so.
  However, this  does not cause any  imprecision in the way  of the GC
  marking more cells as live,  since, during execution, \py\ is yet to
  come  into  existence at  $\pi_6$,  and  its  liveness will  not  be
  consulted by the garbage collector at this point.
\item In an eager language, \py\ would not have been live at $\pi_8$,
  a program point  beyond its last use.  In  a lazy language, however,
  the  definition $\pz  \leftarrow  (1+\py)$, does  not  result in  an
  evaluation  of  $(1+\py)$;  instead   a  closure  is  created.   The
  evaluation  actually takes  place while  evaluating $(\RETURN~\pz)$.
  We therefore  record  the liveness of $\py$ at the program  
  point $\pi_8$,  indeed at every  program point from $\pi_8$  up to
  the beginning  of the program $\pi_1$ (since  we do not  kill liveness).
\end{enumerate}

We thus regard  the body of the function as a  tree and consider paths
from  the  root   of  this  tree  to  program   points  which  trigger
evaluation\footnote{Program   points  which  trigger   evaluation  are
  labelled  with   $\psi$  {\color  {red}  ---slightly   late  in  the
    day}}---\SIF\ conditions  and \RETURN\ expressions.   We call such
paths {\em e-paths}.   For the function \length, there  are three such
paths  $ep_1   =  \langle   \pi_1,  \pi_2, \psi_1\rangle$,   $ep_2  =
\langle\pi_1, \pi_2, \pi_3, \pi_4, \psi_2\rangle$ and  $ep_3 = \langle \pi_1, \pi_2,
\pi_5,  \pi_6, \pi_7, \pi_8, \psi_3\rangle$.

Given  a function  $\mathit{f}$ and  a demand  $\sigma$,  the  liveness
analysis computes two maps:  the function $\mathcal{L}$ computes a map
from a  e-path to a  liveness environment arising  out of all  uses of
variables  along   the  path.  We  call   this  map  $\cal{M}_{\mathit
  f}^{\sigma}$ (or  $\cal{M}_{\mathit f}$ when the  demand $\sigma$ is
clear from  the context).  A  second map $\cal{P}_{\mathit f}$  maps a
program point to  the set of e-paths sharing the  point.  If a program
point happens  to fall on more  than one such path,  then the liveness
environment of  the program  point is the  variable-wise union  of the
liveness  environments  of  the  individual  paths.   In  the  example
program,  since ${\mathcal P}_{\length}(\pi_1)  = \lbrace  ep_1, ep_2,
ep_3\rbrace$,  the liveness  at $\pi_1$  is  ${\cal M_{\length}}(ep_1)
\cup {\cal  M_{\length}}(ep_2)\cup {\cal M_{\length}}(ep_3)$.   In the
sequel   we   shall  describe   how   to   obtain  $\cal{M}$   through
$\mathcal{L}$.  The  formulation of  $\cal{P}$ is easy  and we  do not
elaborate it any further.

@@

The  expression $(\RETURN~\psi:x)$  with a  demand $\sigma$  forces an

@@

is created  and added to the  union.  Also notice  a consequences of
laziness---the  entire  expression  including  the  condition  is  not
